---
title: "Data management examples using the bccc dataset"
author: "Garth Rauscher"
date: "`r Sys.Date()`"
---

Here are some useful resources:


R for applied epidemiology and public health: The Epidemiologist R Handbook
<https://epirhandbook.com/en/index.html>

Each time you start a new R/RStudio session, open and run: 
"EPID501/1.installing and loading packages" first to load all the required packages. Once you have all of the packages loaded, you might ask, "where are all of my installed R packages stored on my hard drive?" 

```{r}

# what is my home or working directory?
R.home(component = "home")
path.expand("~")

# where are all of my installed R packages stored on my hard drive?" 
.libPaths() 

```


## Import your data

The first thing we usually need to do is get your data into a format readable by R, in a dataset known either as a "data frame" or a "tibble" (from the tidyverse package), both having the file extension .Rda.

<https://tibble.tidyverse.org/>

A tibble, or tbl_df, is a modern reimagining of the data.frame, keeping what time has proven to be effective, and throwing out what is not.

We can import the course Dataset into a dataframe (.Rda) from different starting formats, by:
using read.csv (base function) to import utf csv files
using read_dta (haven package) to import stata version 12 or older datasets
using read_sas (haven package) to import SAS  .sas7bdat datasets
using read_sav (haven package) to import SPSS .sav datasets

<https://haven.tidyverse.org/index.html>


You will typically need to read the data into and R dataset each tme you start up R

```{r}

# Let's read the Stata class dataset named bccc.dta (stored in the Data folder) into an R Data Frame

bccc  <- read_dta("data/bccc.dta")


```



```{r}

# There are many things we can do to examine aspects of your data frame:

dim(bccc) # Number of observations and variables
str(bccc) # proc contents

typeof(bccc)                # Data Frames are a type of list in R (not that important to know)
class(bccc)                 # a data frame or tbl or tbl_df )
names(bccc)                 # lists variable names
sapply(bccc, class)         # lists variable names and whether they are numeric or character
sapply(bccc, attributes)    # lists variable names, labels and formats for each variable
attributes(bccc)            # lists dataset attributes (class, notes, row.names, variable names)
summary(bccc)               # proc means
head(bccc)                  # print the first few observations

# You can save the bccc.Rda dataframe so that you can open it in your next session (or you can just import it again from state using read_dta which is what I do).

save(bccc,file="data/bccc.Rda")

```



Two ways you can keep a subset of variables (columns) in your data frame.

In this example, I create a new data frame (bccc2) that contains a subset of the variables from bccc.

```{r}

dim(bccc)  # before

bccc2 <- select(bccc, id,  ethnicity, age, lackprivate, mamhistory ,symptomatic)

dim(bccc2) # after
str(bccc2) # after
summary(bccc2)
head(bccc2)

 
```


```{r}

# OR (another way to create a subset of variables)

# create an object with c() naming all the variables to be kept
dim(bccc)  # before
keeps <- c("id","ethnicity","age","lackprivate","mamhistory","symptomatic")
bccc2<-bccc[ , !(names(bccc) %in% keeps)]
dim(bccc2) # after
str(bccc2) # after

```



Related topic: Use select() to examine distributions and attributes for a subset of variables from your data frame.

<https://sparkbyexamples.com/r-programming/r-select-function-from-dplyr/>


```{r}

    str(select(bccc,"id","ethnicity","age","lackprivate","mamhistory","symptomatic"))  # proc contents; 
summary(select(bccc,"id","ethnicity","age","lackprivate","mamhistory","symptomatic"))  # proc means; 
  print(select(bccc,"id","ethnicity","age","lackprivate","mamhistory","symptomatic"))  # proc print; 

```


How to drop variables from a data frame (rather than keep variables)

First, obtain a list of the variable names for copying and pasting

```{r}
names(bccc) 
```

# "id", "astage", "fnlwt", "age", "age3", "ethnicity", "minority", "incord", "inc", "education", "edu", "disadvantage", "affluence", "muaarea3", "napbc_present", "ncobc_present", "bicoe_present", "ds_present", "rt_recommend", "ct_recommend", "ht_recommend", "insurecat", "mam5yr", "mamhistory", "lastroutine", "lastcbe", "lastmam", "lackprivate", "noregular", "symptomatic", "erprneg", "maxgrade", "pstages", "laterstage", "sumlacktrust", "misconceptions sumcomorbidity", "nlivebirths", "nulliparous", "agefirstbirth", "agelastbirth", "fhxcat", "bmi", "bmicdc_", "deadbc", "blackwhite", "hilowgrade"


Create an object using c() that contains the names of the variables we want to drop I named it "drops".
Then copy the dataframe, excluding the variables listed in "drops" using !names()

```{r }

dim(bccc) # before

drops <- c("astage", "fnlwt", "age3", "minority", "incord", "inc", "education", "edu", "disadvantage", "affluence", "muaarea3", "napbc_present", "ncobc_present", "bicoe_present", "ds_present", "rt_recommend", "ct_recommend", "ht_recommend", "insurecat", "mam5yr", "lastroutine", "lastcbe", "lastmam", "noregular", "erprneg", "maxgrade", "pstages", "laterstage", "sumlacktrust", "misconceptions", "sumcomorbidity", "nlivebirths", "nulliparous", "agefirstbirth", "agelastbirth", "fhxcat", "bmi", "bmicdc_", "deadbc", "blackwhite", "hilowgrade")

bccc2<-bccc[ , !(names(bccc) %in% drops)]

dim(bccc2) # after
str(bccc2) # after

```


Alteratively, you can drop the same variables using select(-c())

```{r }
bccc %>% 
  select(-c(astage, fnlwt, age3, minority, incord, inc, education, edu, disadvantage, affluence, muaarea3, napbc_present, ncobc_present, bicoe_present, ds_present, rt_recommend, ct_recommend, ht_recommend, insurecat, mam5yr, lastroutine, lastcbe, lastmam, noregular, erprneg, maxgrade, pstages, laterstage, sumlacktrust, misconceptions, sumcomorbidity, nlivebirths, nulliparous, agefirstbirth, agelastbirth, fhxcat, bmi, bmicdc_, deadbc, blackwhite, hilowgrade)) %>%   
names()

```


## Reorder variables in your data frame

You can use select() to reorder variables by listing those you want at the front of the dataset first, followed by everything() to tell R to include all the other variables in their original order.

```{r }

str(bccc)  # before

bccc_reorder <- select(bccc, id,  ethnicity, age, lackprivate, mamhistory ,symptomatic)

str(bccc_reorder) # after

```


Rename variables- lets renames incord to the more familiar "income", pstages to the more recognizeable "pathstage", and remove the "_" suffix from the variable name "bmicdc_" 

```{r}

bccc_rename <-
  bccc %>%
  rename(
    income=incord,
    pathstage=pstages,
    bmicdc = bmicdc_
    )

str(bccc_rename)

```



Nominal categorical variables have to be converted or copied to so-called factor variables in order to conduct certain analysis with them. Essentially, converting a variable to a factor creates indicator variables based on the nominal variable.

A nominal categorical variable that is defined as numeric must first be converted to a character variable before it can be converted to a factor variable. In this example, lasmam is a numeric variable and I want to be able to treat is as a factor variables in some analyses.

```{r }
str(bccc$lastmam) # numeric

# convert to character
bccc$lastmam_c <- as.character(bccc$lastmam) 
str(bccc$lastmam_c) # character

# convert to factor
bccc$lastmam_f <- as.factor(bccc$lastmam)
str(bccc$lastmam_f)  # factor
```  




  
  
Drop and keep observations (rows)


Drop observations with specific values of id (1 and 2)

```{r}

dim(bccc)
bccc2<-bccc[!(bccc$id == 1 | bccc$id == 2),]   # uses the OR operator
dim(bccc2)

```


Drop observations with specific values of id (1 and 2), this time by subsetting to all values of id not equal to 1 or 2

```{r}

dim(bccc)
bccc2subset(bccc, bccc$id !=1 & bccc$id !=2)  # uses the AND operator
dim(bccc2)

```


Either method is fine



Keep observations where age >50

```{r}

# before
dim(bccc)
table(bccc$age)
summary(bccc$age)

bccc2<-subset(bccc, bccc$age>50)

# after
dim(bccc2)
table(bccc2$age)
summary(bccc2$age)


```

Keep observations where age <= 50

```{r}


# before
dim(bccc)
table(bccc$age)
summary(bccc$age)

bccc2<-subset(bccc, bccc$age <= 50)

# after
dim(bccc2)
table(bccc2$age)
summary(bccc2$age)

```




Drop variables and observations in sequence, two ways.

The first way involves using select() to drop variables and subset() to drop observations.

```{r }

dim(bccc) 

# subset to age<=50 & education>12       # start with dataframe bccc -> a -> final dataset b
a<-select(bccc, age, education)
b<-subset(a, age <= 50 & education>12)
dim(b)
summary(b)

```

The second way involves piping together the select() and subset() functions with the pipe function %\>%.

```{r }

     
# subset to age<50 & education>12

bccc2<-bccc %>%
  filter(age <= 50 & education>12) %>%
  select(id, ethnicity, age, education)

dim(bccc2)
summary(bccc2)

```




Now, let's keep observations == age 50 & education == 12  

```{r}


dim(bccc)      
table(bccc$age, bccc$education) # before

# subset to age<50 & education>12

bccc2<-bccc %>%
  filter(age == 50 & education == 12) %>%
  select(id, age, education)

dim(bccc2)
table(bccc2$age, bccc2$education) # after




```





## Missing Data and recoding

In this example: values of pstages that represent in-situ stage at diagnosis (pstages=0) are recoded to the value 9.

Recoding a variable using mutate()
Here we recode pstages values of 0 to 9, and leave the other values as is.

```{r}

# recode 0 to 9
table(bccc$pstages) # before
a <- bccc %>% mutate(pstages = if_else(pstages == 0, 9, pstages))

table(a$pstages)    # after

```


Recoding a variable using mutate()
Here we recode values of 0 for pstages to the missing value NA, and leave the other values as is.

```{r}

# recode 0 to NA
b <- bccc %>% mutate(pstages = if_else(pstages == 0, NA, pstages))

table(bccc$pstages) # before
table(b$pstages)    # after

```

## Recoding invalid values to missing values (NA)

The three variables **lastroutine, lastmam, lastcbe** in the data frame "bccc" have their values of 4 and 9 recoded to NA

```{r }

summary(select(bccc,lastroutine,lastmam,lastcbe))  # before recoding

unique(bccc$lastroutine)
unique(bccc$lastmam)
unique(bccc$lastcbe)

# recode 4 and 9 to NA
bccc2<-bccc # copy original dataset into bccc2 and keep using bccc2 throughout this block of code
bccc2 <-  mutate_at(bccc2, vars(lastroutine,lastmam,lastcbe), ~ na_if(.,9))
bccc2 <-  mutate_at(bccc2, vars(lastroutine,lastmam,lastcbe), ~ na_if(.,4))


# Three different ways to check the recoded values

table(select(bccc2,lastroutine,lastmam,lastcbe)) # after

# after

table(bccc2$lastroutine)
table(bccc2$lastmam) 
table(bccc2$lastcbe) 

unique(bccc2$lastroutine)
unique(bccc2$lastmam)
unique(bccc2$lastcbe)



```



More examples using the dplyr package

https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html

```{r }
bccc  # a tibble 989 × 50

```

Filter or select observations (rows)
```{r }
# before
table(bccc$ethnicity)
table(bccc$lackprivate) 


a<-bccc %>% filter(ethnicity == "1", lackprivate=="0")

# after
table(a$ethnicity)
table(a$lackprivate) 



```


Sorting rows with arrange()

```{r }
a<-bccc %>% arrange(ethnicity, age)
a
```


Select columns with select()

```{r }
a<-bccc %>% select(ethnicity, age)
str(a)
```


Select and rename variables

```{r }
a<-bccc %>% select(pathstage=pstages, income=incord)
str(a)
```


Rename variables without selecting

```{r }
a<-bccc %>% rename(pathstage=pstages, income=incord)
str(a)
```


Change column order with relocate()

```{r }
a<-bccc %>% relocate(deadbc)
str(a)
```


You can add a column using mutate:

```{r }

# create a variable for decade of age
a <-bccc %>% mutate(decade=age/10) 
a$decade <-floor(a$decade)

# run error check to compare old to new values, using the freqtables package
table(a$decade,a$age)

```




## Define a categorical variable from a continuous variable

```{r}

bccc2 <- 
  mutate(
    bccc,
    age3 = case_when(
      age >= 18 & age<=49 ~ 1,
      age >= 50 & age<=64 ~ 2,
      age >= 65 & age<=80 ~ 3,
      TRUE ~ NA
      )
    )

# run error check to compare old to new values, using freqtables package
bccc2 %>%
freq_table(age3, age) %>%
freq_format(recipe = "n", name = NA, digits = NA) %>%
select(1,2,3) # select the bare minimum for checking purposes.

```


Standardize a variable to have a mean of 0 ad a SD of 1

```{r}
# age normalized to mean of ALL rows - use the mean and sd fuctions of base R

bccc$age_z= (bccc$age-mean(bccc$age)) /sd(bccc$age)

mean(bccc$age_z)
sd(bccc$age_z)

```

Another way to check the distribution of your new variable using the procs package

```{r}

# check result (we will review the procs package later)

proc_means(bccc,
  var = v(age,age_z),
  stats = c("n", "mean", "std", "min", "max"),
  output = NULL,
  by = NULL,
  class = NULL,
  options = NULL,
  titles = NULL
)
```


How to count and list distinct values for a variable

There are a number of ways to do this. The two methods below take the variable age3 from the data frame bccc and counts/lists the number of distinct values.

```{r }

# list distinct values
sqldf("select distinct(age3) from bccc")

# number of distinct values
bccc %>% summarise(distinct_values = n_distinct(age3))


```


## Sorting and merging

Here I subset bccc into 2 data frames (**a** and **b**) with different variables then merge the variables back together on "id" (this variabe must be in oth datasets for the merge to work).

```{r}

a<-select(bccc,id,age,age3,ethnicity)
b<-select(bccc,id,edu,inc)


# sort on the by variable id in increasing value (default is decreasing)
a<-(a[order(a$id, decreasing = FALSE), ])
b<-(b[order(b$id, decreasing = FALSE), ])

# merge on the by variable GEOID
c <- merge(a, b, by.x = "id", by.y = "id") # by.x refers to 1st dataframe listed, by.y to the second one.

str(c)

```


How to check if there are any duplicates on all variables in your dataset

```{r}

bccc %>%  distinct()  #987 distinct-no duplicate observations

```


De-duplication of observations with identical values on specific variables.
In this example I want to de-duplicate by age. 
This is just an example.
I create the object "nodups_by_age" (I could have named it "snoopy") and use distinct()

```{r}

# keep one observation for each value of age
nodups_by_age<-bccc %>% distinct(age, .keep_all = TRUE)

dim(nodups_by_age)  # 54 unique values of age
table(nodups_by_age$age) # 1 onbs for each value of age

```


Another way to do this

```{r}

# keep one observation for each value of age
nodups_by_age<-bccc %>% distinct(age, .keep_all = TRUE)


# two ways to check your result

sqldf("select distinct(age) from nodups_by_age")

table(nodups_by_age$age)

# how many distinct values of age are there?
bccc %>% summarise(distinct_values = n_distinct(age)) # Answer: 54

```


