---
title: "Standardization"
author: "Rauscher, Garth H"
date: "`r Sys.Date()`"
output: html_document
---


R for applied epidemiology and public health: The Epidemiologist R Handbook
<https://epirhandbook.com/en/index.html>

Getting started in R for Stata users
<https://dss.princeton.edu/training/RStata.pdf>



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

R.Version()
```


Import your data

The first thing we usually need to do is get your data into a format readable by R, known as a data frame and having the extension .Rda.

Below we import the course Stata Dataset into R using **read_dta** (using the **haven** package)



```{r }

# read th Stata dataset named bccc.dta (stored in the Data folder) into an R Data Frame
bccc <- read_dta("data/bccc.dta")

# Number of observations and variables
dim(bccsv)

# proc contents
str(bccc)

# proc means
summary(bccc)

# print the first few observations
head(bccc)


# save the bccc.Rda dataframe
save(bccc,file="data/bccc.Rda")


```
 


Stratified tables with counts and percentages and tests but no pooled measures of association 


janitor package for stratified 2x2 tables (counts only)

tabyl(exposure, outcome, stratification variable)

like proc freq tables stratavar*exposure*outcome

```{r}

# stratified 2x2 tables
bccc %>% tabyl(lackprivate, symptomatic,age3)   

# stratified 2x2 tables, adorned
bccc %>% tabyl(lackprivate, symptomatic,age3) %>%  
    adorn_percentages(denominator = "row")    %>% 
    adorn_pct_formatting(digits = 0)          %>% 
    adorn_title(row_name = "lackprivate",col_name = "symptomatic")

# Chi-squared test of association
age_by_outcome <- bccc  %>% 
  tabyl(age3, symptomatic, show_na = FALSE) 
chisq.test(age_by_outcome)

```



 
The risks package
Stratified tables with counts and percentages and tests AND pooled measures neasures for RD and RR !!!!!
https://rdrr.io/github/stopsack/risks/man/rr_rd_mantel_haenszel.html

 
```{r}
library(risks)

bccc2<-dplyr::select(bccc, age3, lackprivate, symptomatic)

rr_rd_mantel_haenszel( bccc2, lackprivate, symptomatic, age3, estimand=c("rr"), conf.level=0.95 )
rr_rd_mantel_haenszel( bccc2, lackprivate, symptomatic, age3, estimand=c("rd"), conf.level=0.95 )

```



epi.2by2

Computes summary measures of risk and a chi-squared test for difference in the observed proportions from count data presented in a 2by2 table. With multiplevstrata the function returns crude  and Mantel-Haenszel adjusted measures of association and chi-squared tests of homogeneity.

<https://rdrr.io/cran/epiR/>
<https://search.r-project.org/CRAN/refmans/epiR/html/epi.2by2.html>
<https://github.com/cran/epiR/blob/master/R/epi.2by2.R>

epi.2by2

Computes several measures of association

Arguments
dat        - a vector of length four, an object of class table or an object of class grouped_df from package dplyr 
             containing the individual cell frequencies (see below).
method     - a character string indicating the study design on which the tabular data has been based. Options are 
             cohort.count, cohort.time, case.control, or cross.sectional. Based on the study design specified by the user,  
             appropriate measures of association, measures of effect in the exposed and measures of effect in the population 
             are returned by the function.
digits     - scalar, number of digits to be reported for print output. Must be an integer of either 2, 3 or 4.
conf.level - magnitude of the returned confidence intervals. Must be a single number between 0 and 1.
units      - multiplier for prevalence and incidence (risk or rate) estimates.
interpret  - logical. If TRUE interpretive statements are appended to the printepi.2by2 object.
outcome    - a character string indicating how the outcome variable is represented in the contingency table. Options   
             are as.columns (outcome as columns) or as.rows (outcome as rows).
x, object  - an object of class epi.2by2.


Unfortunately I could not get epi.2by2 to run because it could not locate the dependency, "zexact"

```{r }
library(epiR)
library(spatstat.geom)

dim(bccc)


df<-bccc %>% filter(ethnicity==1) 

table(df$lackprivate, df$symptomatic) #obtain a,b,c,d


## Lackprivate=1 (non, sym) 24,14  (a,b)
## Lackprivate=0 (non, sym) 137,222 (c,d) 

## Outcome variable (FUS) as columns:
dat <- c(24,14,137,222) # a,b,c,d
# View the data in the usual 2 by 2 table format:
matrix(dat, nrow = 2, byrow = TRUE)

epi.2by2(dat = dat, method = "cross.sectional", conf.level = 0.95, units = 1, 
   interpret = FALSE, outcome = "as.columns")

epi.2by2(dat = dat, method = "cohort.count", conf.level = 0.95, units = 1, 
   interpret = FALSE, outcome = "as.columns")

 
```

