---
title: "Models"
author: "Garth Rauscher"
format: html
editor: source
---

R for applied epidemiology and public health: The Epidemiologist R Handbook
<https://epirhandbook.com/en/index.html>



```{r}

# read th Stata dataset named bccc.dta (stored in the Data folder) into an R Data Frame
bccc <- read_dta("data/bccc.dta")

```



#### FIRST PRIZE FOR CONDUCTIG LINEAR REGRESSIONS !!!!!

#### Ordinary Least Squares (Method #2)

This method takes advantage of the lm_robust function with the "classical" standard error calculation to produce ready made output for ordiary least squares. <https://search.r-project.org/CRAN/refmans/estimatr/html/lm_robust.html> <https://www.rdocumentation.org/packages/estimatr/versions/0.30.6/topics/lm_robust>

```{r }

# Ordinary Least Squares using lm_robust (se_type =  "classical")

library(estimatr)

mod2 <-
  lm_robust(symptomatic~lackprivate+age,
  data=bccc,
  subset=ethnicity==1,
  weights=fnlwt,
  clusters=NULL,
  se_type =  "classical",
  ci = TRUE,
  alpha = 0.05,
  return_vcov = TRUE,
  try_cholesky = FALSE
)
mod2

#str(mod2) # gives you all the information from the model, similar to Stata's ereturn list

```




Modified Ordinary Least Squares with robust standard errors (HC3) for conducting linear risk difference models.

This method uses the **lm_robust** function of the **estimatr** package

<https://www.rdocumentation.org/packages/estimatr/versions/0.30.6/topics/lm_robust>

```{r }

# Ordinary Least Squares with Robust Standard Errors (se_type="HC3")

library(estimatr)

mols1 <-
  lm_robust(symptomatic~lackprivate + age,
  data=bccc,
  subset=ethnicity==1,
  weights=fnlwt,   # ths example used the fnlwt variable containing weights
  clusters=NULL,
  se_type =  "HC3",
  ci = TRUE,
  alpha = 0.05,
  return_vcov = TRUE,
  try_cholesky = FALSE
)
mols1


# str(mols1)

```







Standard Poisson Regression

```{r }

# model equation
fun<-symptomatic~lackprivate+age

# model equation nested in glm function
fish1 <- glm(fun, 
             data    = bccc,
             family  = poisson(link="log"),
             subset  = ethnicity==1,
             weights = fnlwt
)
fish1
```



Extract, calculate and paste together results using 95% *****Wald-based***** CI

Advantage: get all the estimates

Disadvantage: 95% CI not quite as good those from profile-based methods, although
we do routinely use wald-based 95% CI


```{r }

results <- coeftest(fish1)


results[,1]  # coefficients
results[,2]  # standard errors

# create a table with your expnentiated coefficientss, confidence intervals, and p-values
model_results <- round(cbind((cbind(RR  = exp(results[,1]), 
                                       LCI = exp(results[,1] + qnorm(0.05/2)*results[,2]),
                                       UCI = exp(results[,1] - qnorm(0.05/2)*results[,2]))),
                                       P = results[,4]),3)
model_results


```





Extract, calculate and paste together results using 95% *****profile-based CI*****
The confint function return profile-based intervals.

Advantage: Profile-based 95% CI can be better than wald intervals
Disadvantage: with this method you only get one at a time

```{r }

# extract point estimate
point <- exp(fish1$coefficients)[2]

# calculate confidence interval
ci<-exp(confint(fish1, 'lackprivate', level=0.95))

# paste them together and print
result<-paste(c(point, ci))

# convert the list to a numeric vector
v <- as.vector(unlist(result),'numeric')
options(digits=4)
print(v)

```


Modified Poisson Regression with robust standard errors for conducting  risk ratio models.

This method uses the glm funtion along with coeftest to calculate 95% CI based on a sandwich estimate of the SE.
Point estimates and confidence bounds are then pasted together to create the results.

```{r }

# model equation
fun<-symptomatic~lackprivate+age

# model equation nested in glm function
mod4 <- glm(fun, 
             data    = bccc,
             family  = poisson(link="log"),
             subset  = ethnicity==1,
             weights = NULL
)
mod4
results <- coeftest(mod4, vcov = sandwich)

# create a table with your exponentiated coefficients, confidence intervals, and p-values
model_results <- round(cbind(exp(cbind(RR  = results[,1], 
                                       LCI = results[,1] + qnorm(0.05/2)*results[,2],
                                       UCI = results[,1] - qnorm(0.05/2)*results[,2])),
                                       P = results[,4]),3)
options(digits=4)
print(model_results)


```





Logistic Regression

In this example we are subsetting to NH White patients

```{r }

library(lmtest)
library(sandwich)
library(tidyverse)

bccc2<-select(bccc,symptomatic,lackprivate,age,lastmam, lastcbe, lastroutine, ethnicity) %>% filter(ethnicity==1, na.rm = TRUE)

# model equation
fun<-symptomatic~lackprivate+age+lastmam+lastroutine+lastcbe

# model equation nested in glm function
mod4 <- glm(fun, 
             data    = bccc2,
             family  = binomial(link="logit"),
             subset  = ethnicity==1,
             #weights = fnlwt
)
mod4
results <- coeftest(mod4)

round(results[,1],3)  # coefficients
results[,2]  # standard errors

# create a table with your expnentiated coefficientss, confidence intervals, and p-values
model_results <- round(cbind(exp(cbind(OR  = results[,1], 
                                       LCI = results[,1] + qnorm(0.05/2)*results[,2],
                                       UCI = results[,1] - qnorm(0.05/2)*results[,2])),
                                       P = results[,4]),3)
options(digits=4)
model_results

```


Let's turn these ORs into something more useful !!!!!


Conditional standardizated point estimates for R1, R0 and RD

Conditional standardization estmates the point and interval estimates at a specific combination of covariate values (e.g., and mean or median of all covariates).

Here is an example applying conditional standardization to the logistic regression model results object (mod4). COnditionally standardized at the mean of binary/ordinal variables and at the median of continuous age.

Note: This method does not yield confidence intervals, just point estimates.

```{r }

mod4  # among NHW patients

bccc2<-na.omit(bccc2) # have to remove NAs from the variables being used

lastmam_scal     <- mean(bccc2$lastmam)
lastcbe_scal     <- mean(bccc2$lastcbe)
lastroutine_scal <- mean(bccc2$lastroutine)
age_scal         <- median(bccc2$age)

lastmam_scal
lastcbe_scal
lastroutine_scal
age_scal


# predicted probablity of a symotomatic detection for someone without private insurance (index), conditionally (among those who are age 50, NH White, with a recent mammogram, a recent cbe and a recent routine physical)

P1 <-predict(mod4, newdata = data.frame(lackprivate = 1, ethnicity=1,lastmam=lastmam_scal, lastcbe=lastcbe_scal, lastroutine=lastroutine_scal, age = age_scal),type="response")

P1

# predicted probablity of a symotomatic detection for someone with private insurance (referent), conditionally (among those who are age 50, NH White, with a recent mammogram, a recent cbe and a recent routine physical)

P0 <-predict(mod4, newdata = data.frame(lackprivate = 0, ethnicity=1,lastmam=lastmam_scal, lastcbe=lastcbe_scal, lastroutine=lastroutine_scal, age = age_scal),type="response")


P0



# predicted prevalence difference in symptomatic detection comparing someone without (index)  v. with (referent) private insurance, conditionally (among those who are age 50, NH White, with a recent mammogram, a recent cbe and a recent routine physical)

PD = P1-P0
PR = P1/P0
OR = (P1/(1-P1))/(P0/(1-P0))

model_results <- round(cbind(P1, P0, PD, PR, OR),3 )
model_results[1,]

```



Marginal standardization point estimates, confidence intervals and p-values

An Introduction to ‘margins’

https://cran.r-project.org/web/packages/margins/vignettes/Introduction.html
https://cran.r-project.org/web/packages/margins/vignettes/Introduction.html#Average_Marginal_Effects_and_Average_Partial_Effects

Calculate the marginal standardized prevalence difference for the association of lackprivate with symptomatic detection among NH WHite patients, adjusted for age, lastmam, lastcbe and lastroutine. NOTE: this is just an example and these results may not be of interest to anyone.

```{r }


# create a dataset with variables of interest and among NH White patients only for this example, and remove observations with missing values.

bccc2<-select(bccc,symptomatic,lackprivate,age,lastmam, lastcbe, lastroutine, ethnicity) %>% filter(ethnicity==1, na.rm = TRUE)

# model equation
fun<-symptomatic~lackprivate+age+lastmam+lastroutine+lastcbe

# logistic regression
# model equation nested in glm function
mod5 <- glm(fun, 
             data    = bccc2,
             family  = binomial(link="logit"),
             subset  = ethnicity==1,
             #weights = fnlwt
)
mod5

margins(mod5) # point estimates
summary(margins(mod5, variables = "lackprivate")) # RD (95% CI)

```



Using the at Argument

The at argument allows you to calculate marginal effects at representative cases (sometimes “MERs”) or marginal effects at means - or any other statistic - (sometimes “MEMs”), which are marginal effects for particularly interesting (sets of) observations in a dataset.

This differs from marginal effects on subsets of the original data in that it operates on a modified set of the full dataset wherein particular variables have been replaced by specified values. This is helpful because it allows for calculation of marginal effects for counterfactual datasets (e.g., what if all women were instead men? what if all democracies were instead autocracies? what if all foreign cars were instead domestic?).


```{r }
mod5

# RD due to lackprivate across different values of age (adjusted for the other covariates)
summary(margins(mod5,variables = "lackprivate", at = list(age=50:59)))
        
# RD due to lackprivate across different values of age (adjusted for the other covariates)
summary(margins(mod5,variables = "lackprivate", at = list(age=seq(20, 80, by=10))))

```













